---
title: "Baseball Analysis - Tara Ghorpadkar"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

Motivation:
I wanted to analyze the trends in baseball following the recent enforcement of the "sticky stuff" ball rule change in the MLB implemented on June 21, 2021. This rule prohibits the doctoring of balls by the pitching team in order to increase the spin rate by getting a better grip on the ball. By increasing the spin rate with the use of adhesives on the ball, this increases the ball's time in the air, therefore making the ball's vertical position higher than expected when it reaches the hitter. This rule was created prior to June 21, but it had been loosely enforced until recently. Suspicions of teams using sticky substances on baseballs arose back in 2018 when The Dodgers' Trevor Bauer blew the whistle on the league, but enforcement continued to be loose until this past season saw the lowest batting averages and the highest no-hitters on record. The MLB came out with the news on June 5, 2021 to warn the teams of this upcoming change. This new rule will suspend a player for 10 days if they are caught using sticky substances to enhance their performance.For my project, I wanted to see how and when the rule change affected the league. Specifically, I looked to see which types of pitches were most impacted by the new rule, when exactly the change in spin rate occurred, which teams previously relied on the use of sticky substances the most, and whether batting averages were impacted by this enforcement.

```{r message=FALSE, warning=FALSE}
library(devtools)
library(baseballr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(scales)
library(reshape2)
library(knitr) 


```

First, I import the data from Baseball Savant Statcast Search from the beginning to July 30 of the 2021 season. In order to download the data from the website, you will need to uncomment the lines in this chunk.
```{r message=FALSE, warning=FALSE}
#dates <- seq.Date(as.Date('2021-04-01'), as.Date('2021-07-30'), by = 7)

#date_grid <- tibble(start_date = dates, end_date = dates+6)

#savant_data <- purrr::map2_df(.x = date_grid$start_date,
#                              .y = date_grid$end_date,
#                              ~scrape_statcast_savant(start_date = .x,
#                                                      end_date = .y,
#                                                      player_type = 'pitcher'))
#write.csv(savant_data,"savant_data.csv", row.names = TRUE)

df = read.csv("savant_data.csv", header = TRUE)
paste('Number of pitches since July 30:', nrow(df))

```

Then I clean the data frame by removing all of the columns that I am not using for my analysis as well as null values for the columns I am using.
```{r}
df[,c("pitch_type", "game_date", "release_speed",
      "release_pos_x", "release_pos_z", "player_name",
      "zone", "des", "p_throws", "home_team", "away_team",
      "plate_x", "plate_z", "inning_topbot", "launch_speed",
      "launch_angle", "effective_speed", "release_spin_rate",
      "release_extension", "estimated_ba_using_speedangle",
      "estimated_woba_using_speedangle", "woba_value", "woba_denom",
      "babip_value", "iso_value", "launch_speed_angle",
      "at_bat_number", "pitch_number", "pitch_name", "spin_axis")]

# remove NAs
df <- df[!(is.na(df$release_spin_rate)|is.na(df$pitch_type)|is.na(df$pitch_name)|is.na(df$zone)), ]

```

To normalize the spin rates in order to compare spin rates with different pitch velocities, I convert the spin rates to Bauer Units in a Bauer Units Column. I also convert the date values to Date data types in R. I then designate which team is pitching. I also remove pitch types not commonly used in the league.
```{r}
#create Baur units column
df$b_units <- df$release_spin_rate/df$release_speed

#month and week of game date
df$month <- month(as_date(df$game_date))
df$week <- week(as_date(df$game_date))
df$week_of <- as.Date(sapply (as_date(df$game_date), function(d) { return (d + (-6 - as.POSIXlt(d)$wday %% -7 ))}), origin = "1970-01-01")
df$pitch_team <- ifelse(df$inning_topbot == "Bot", df$away_team,df$home_team)
df2 <- df[df$pitch_name != "Fastball" & df$pitch_name != "Knuckleball" & df$pitch_name != "Eephus" & df$pitch_name != "2-Seam Fastball" & df$pitch_name != "Screwball",]
```


Here, I have grouped the data by its average daily, weekly, and monthly spin rates and taken averages of the points of data to get one point for each day, week, and month. Then, I have put that data in a table.
```{r}
# start of analysis

daily_avg_spin <- aggregate(df2$release_spin_rate, list(df2$game_date), FUN=mean)
colnames(daily_avg_spin) <- c('date','spin_rate')
weekly_avg_spin <- aggregate(df2$release_spin_rate, list(df2$week_of), FUN=mean)
colnames(weekly_avg_spin) <- c('week_of','spin_rate')
monthly_avg_spin <- aggregate(df2$release_spin_rate, list(df2$month), FUN=mean)
colnames(monthly_avg_spin) <- c('month','spin_rate')
```

These are the graphs of the average daily, weekly, and monthly spin rates from April 1 to July 30.
```{r}
ggplot(data=monthly_avg_spin, aes(x=as.Date(paste0("2021-", month, "-1")), y=spin_rate, group=1)) +
  geom_line(linetype="solid", size=2, color="#0099f9")+
  labs(
    x = "Month",
    y = "Average Spin Rate",
    title = "Average Spin Rate Per Month",
    subtitle = "All Pitches (2021)",
    caption = "Source: Baseball Savant"
  )+
  scale_x_date(date_labels = "%b")
```

```{r}
ggplot(data=weekly_avg_spin, aes(x=week_of, y=spin_rate, group=1)) +
  geom_line(linetype="solid", size=2, color="#0099f9")+
  labs(
    x = "Week Of",
    y = "Average Spin Rate",
    title = "Average Spin Rate Per Week",
    subtitle = "All Pitches (2021)",
    caption = "Source: Baseball Savant"
  ) +
  scale_x_date(breaks = scales::breaks_pretty(10))
```

```{r}
ggplot(data=daily_avg_spin, aes(x=as.Date(date), y=spin_rate, group=1)) +
  geom_line(linetype="solid", size=1, color="#0099f9")+
  labs(
    x = "Date",
    y = "Average Spin Rate",
    title = "Average Spin Rate Per Day",
    subtitle = "All Pitches (2021)",
    caption = "Source: Baseball Savant"
  ) +
  scale_x_date(date_labels = "%b-%e") +
  geom_vline(xintercept = as.numeric(ymd("2021-06-05")), linetype="solid", color = "yellow", size=1.5) +
  geom_text(aes(x=as.Date("2021-06-05"), label="Threat Of Enforcement", y=2340)) +
  geom_vline(xintercept = as.numeric(ymd("2021-06-21")), linetype="solid", color = "red", size=1.5) +
  geom_text(aes(x=as.Date("2021-06-21")+10, label="Enforcement", y=2300))
```
For the graph of the Average Daily Spin Rate, I included a line to mark the date when the threat of enforcement was announced (June 5) as well as a line to mark the date of actual enforcement (June 21). It seems as if the change in spin rate took a dive shortly after threat of enforcement, so I created a graph and a table to get a specific date that accounts for the drop.
```{r}
june_1_15_daily_spin <- daily_avg_spin[daily_avg_spin$date >= '2021-05-31' & daily_avg_spin$date <= '2021-06-15',]

june_1_15_daily_spin

ggplot(data=june_1_15_daily_spin, aes(x=as.Date(date), y=spin_rate, group=1)) +
  geom_line(linetype="solid", size=2, color="#0099f9")+
  labs(
    x = "Date",
    y = "Average Spin Rate",
    title = "Average Spin Rate Per Day",
    subtitle = "All Pitches (2021)",
    caption = "Source: Baseball Savant"
  ) +
  scale_x_date(date_labels = "%b-%e",date_breaks  ="1 day") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The day on which there is a significant drop is June 8th, so up till June 7th, the threat of enforcement did not seem to have an effect on spin rate. The drop, however, remained constant after the June 21 enforcement. This shows that spin rate significantly decreased just by threat of enforcement, which means that teams most likely stopped using sticky substances after they were threatened with a suspension and did not wait until actual enforcement to do so.

It is also clear that universally, pitch spin rates were decreased by this threat. Next, I want to check whether this change in spin rate was actually due to pitchers having a lower spin value regardless of pitch or was it due to the pitchers using pitches that naturally do not have as high of a spin rate. To check this, I am going to visualize the frequency of types of pitches being thrown before and after June 7th and the change in spin rate on each of those pitches.

Now, I am grouping the data by its pitch name and finding the frequency of each pitch being thrown before and after June 7th. I am also finding the change in frequency in each pitch.
```{r}
pre_freq <- df2[df2$game_date <= '2021-06-07',]
pre_freq <- aggregate(pre_freq$release_spin_rate, list(pre_freq$pitch_name), FUN = length)
colnames(pre_freq) <- c('pitch_name','pre_count')

pre_freq$pre_frequency <- round(pre_freq$pre_count / sum(pre_freq$pre_count) * 100, digits = 2)

post_freq <- df2[df2$game_date > '2021-06-07',]
post_freq <- aggregate(post_freq$release_spin_rate, list(post_freq$pitch_name), FUN = length)
colnames(post_freq) <- c('pitch_name', 'post_count')

post_freq$post_frequency <- round(post_freq$post_count / sum(post_freq$post_count) * 100, digits = 2)

pre_post_freq <- merge(pre_freq, post_freq, by = "pitch_name")

pre_post_freq$freq_diff <- pre_post_freq$post_frequency - pre_post_freq$pre_frequency

pre_post_freq$freq_diff_pct_freq <- round((pre_post_freq$freq_diff / pre_post_freq$pre_frequency) * 100, digits = 2)

pre_post_freq
```
As shown, the largest drop in frequency for a pitch was for the Knuckle-Curve pitch.

Next, I am going to look at the spin rates for each pitch before and after June 7th.
```{r}
pre_spin <- df2[df2$game_date <= '2021-06-07',]
pre_spin <- aggregate(pre_spin$release_spin_rate, list(pre_spin$pitch_name), FUN = mean)
colnames(pre_spin) <- c('pitch_name','pre_spin')

post_spin <- df2[df2$game_date > '2021-06-07',]
post_spin <- aggregate(post_spin$release_spin_rate, list(post_spin$pitch_name), FUN = mean)
colnames(post_spin) <- c('pitch_name','post_spin')

pre_post_spin <- merge(pre_spin, post_spin, by = "pitch_name")

pre_post_spin$spin_diff <- pre_post_spin$post_spin - pre_post_spin$pre_spin

pre_post_spin$spin_diff_pct_spin <- (pre_post_spin$spin_diff / pre_post_spin$pre_spin) * 100

pre_post_spin

```
For each pitch, there is a 2.5 - 3.5 % decrease in the spin rate after June 7th, so there was no dramatic decrease in spin rate for the Knuckle-Curve Pitch. It appears as though no matter the pitch, the enforcement of the "sticky stuff" rule decreased spin rate around the same amount. The exception for this is the Split-Finger pitch, but because it has a much lower frequency than any other pitch on the table, it is susceptible to outliers in its data set because the sample size is so small, so this data may not be entirely reliable for that pitch.

Next, I wanted to see which teams were impacted the most by this rule change. Did this problem affect some teams in particular, or was it fairly even across the board?

Here, I am grouping the data by pitch team to find the differences in average spin rate per team before and after June 7th. I have ordered the data in terms of largest to smallest difference.

```{r}
team_pre_spin <- df2[df2$game_date <= '2021-06-07',]
team_pre_spin <- aggregate(team_pre_spin$release_spin_rate, list(team_pre_spin$pitch_team), FUN = mean)
colnames(team_pre_spin) <- c('pitch_team','pre_spin')

team_post_spin <- df2[df2$game_date > '2021-06-07',]
team_post_spin <- aggregate(team_post_spin$release_spin_rate, list(team_post_spin$pitch_team), FUN = mean)
colnames(team_post_spin) <- c('pitch_team','post_spin')

team_spin_comp <- merge(team_pre_spin, team_post_spin, by = "pitch_team")

team_spin_comp$difference <- team_spin_comp$post_spin - team_spin_comp$pre_spin

team_spin_comp <- team_spin_comp[order(team_spin_comp$difference),]
team_spin_comp
```
As shown, the LA Dodgers and the Detroit Tigers have seen the largest drop in average spin rate, but the Chicago White Sox and the Minnesota Twins have not seen a noticable drop in spin rate. Therefore, this rule change does not affect the league evenly, as some teams may have relied more on sticky substances than others and with the new enforecement have seen drastic drops in spin rate as a result.

```{r}
ggplot(team_spin_comp, aes(x = pitch_team, y = difference, group = 1)) +
  geom_line(linetype="solid", size=2, color="#0099f9")+
  labs(
    x = "Pitch Team",
    y = "Spin Rate Difference",
    title = "Difference in Spin Rate by Team",
    subtitle = "All Pitches (2021)",
    caption = "Source: Baseball Savant"
  ) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

It is obvious that this rule has had an affect on teams in the league, but has this rule affected specific players? Next, I want to check which pitchers have seen the largest drops in spin rate as a result of the rule. I group the data by pitcher and average their spin rate before and after June  7th and calculate the difference in spin  rate.
```{r}
pitcher_pre_spin <- df2[df2$game_date <= '2021-06-07',]
pitcher_pre_spin <- pitcher_pre_spin <- aggregate(pitcher_pre_spin$release_spin_rate, list(pitcher_pre_spin$player_name), FUN = mean)
colnames(pitcher_pre_spin) <- c('player_name','pre_spin')

pitcher_post_spin <- df2[df2$game_date > '2021-06-07',]
pitcher_post_spin <- pitcher_post_spin <- aggregate(pitcher_post_spin$release_spin_rate, list(pitcher_post_spin$player_name), FUN = mean)
colnames(pitcher_post_spin) <- c('player_name','post_spin')

pitcher_spin_comp <- merge(pitcher_pre_spin, pitcher_post_spin, by = "player_name")

pitcher_spin_comp$spin_diff <- pitcher_spin_comp$post_spin - pitcher_spin_comp$pre_spin

pitcher_spin_comp <- pitcher_spin_comp[order(pitcher_spin_comp$spin_diff),]
pitcher_spin_comp


```
I also create a table to show the number of pitches thrown by each player.
```{r}
pitches_thrown <- df2
pitches_thrown <- pitches_thrown <- aggregate(pitches_thrown$pitch_name, list(pitches_thrown$player_name), FUN = length)
colnames(pitches_thrown) <- c('player_name', 'pitches_thrown')
```


Here, I create a table to show the number of appearances of each player in games before and after June 7th and the total appearances up till July 30th.
```{r}
df_after_june_7 <- df2[df2$game_date >= '2021-06-07',]
apps_after_grouper <- aggregate(df_after_june_7$game_date, list(df_after_june_7$player_name), FUN = function(x) length(unique(x)))
colnames(apps_after_grouper) <- c('player_name', 'appearances_june_7_later')

df_before_june_7 <- df2[df2$game_date < '2021-06-07',]
apps_before_grouper <- aggregate(df_before_june_7$game_date, list(df_before_june_7$player_name), FUN = function(x) length(unique(x)))
colnames(apps_before_grouper) <- c('player_name', 'appearances_june_7_before')

apps <- merge(apps_after_grouper, apps_before_grouper, by = "player_name")
apps$total_apps <- apps$appearances_june_7_later + apps$appearances_june_7_before


```
This table shows the combined number of pitches and appearances made by each pitcher.
```{r}
pitches_apps <- merge(apps, pitches_thrown, by = "player_name")
pitches_apps$pitches_per_app <- round((pitches_apps$pitches_thrown / pitches_apps$total_apps),digits = 1)
```

This data frame combines the number of pitches, the number of appearances, and the spin rate of each player before and after June 7th.
```{r}
apps_spin_diff <- merge(pitches_apps,pitcher_spin_comp,by = "player_name")
apps_spin_diff
```

Next, I am going to compare the xWOBA and the Bauer Units(pitch spin rate / pitch spin velocity) for each player before and after June 7th.

First, I am going to focus on xWOBA. This is the expected value of a batted ball for a player based on a launch angle and launch velocity when it leaves the bat. Here, I group the data based on player name and find the average xWOBA for each player before and after June 7th to find the difference in xWOBA based on pitcher appearances.
```{r}
pitcher_pre_xwoba <- aggregate(df_before_june_7$estimated_woba_using_speedangle, list(df_before_june_7$player_name), FUN = mean, na.rm=TRUE, na.action=na.pass)
colnames(pitcher_pre_xwoba) <- c('player_name', 'Pre_xwOBA')
pitcher_post_xwoba <- aggregate(df_after_june_7$estimated_woba_using_speedangle, list(df_after_june_7$player_name), FUN = mean, na.rm=TRUE, na.action=na.pass)
colnames(pitcher_post_xwoba) <- c('player_name', 'Post_xwOBA')
pitcher_xwoba_comp <- merge(pitcher_pre_xwoba,pitcher_post_xwoba,by = "player_name")
pitcher_xwoba_comp$xwOBA_diff <- pitcher_xwoba_comp$Post_xwOBA - pitcher_xwoba_comp$Pre_xwOBA
apps_spin_xwoba <- merge(apps_spin_diff, pitcher_xwoba_comp, by = "player_name")
apps_spin_xwoba
```

Second, I am going to focus on Bauer Units. Here, I group the data based on player name and average the Bauer Units before and after June 7th to find the difference in Bauer Units due to the rule change.
```{r}
pitcher_pre_bu <- aggregate(df_before_june_7$b_units, list(df_before_june_7$player_name), FUN = mean)
colnames(pitcher_pre_bu) <- c('player_name', 'Pre_b_units')
pitcher_post_bu <- aggregate(df_after_june_7$b_units, list(df_after_june_7$player_name), FUN = mean)
colnames(pitcher_post_bu) <- c('player_name', 'Post_b_units')
pitcher_bu_comp <- merge(pitcher_pre_bu, pitcher_post_bu, by = "player_name")
pitcher_bu_comp$bu_diff <- pitcher_bu_comp$Post_b_units - pitcher_bu_comp$Pre_b_units
pitcher_bu_comp
```

Now I merge the xWOBA comparison data frame with the Bauer Units comparison data frame.
```{r}
apps_spin_xwoba_bu <- merge(apps_spin_xwoba,pitcher_bu_comp, by = "player_name")
apps_spin_xwoba_bu$percent_spin_change <- (apps_spin_xwoba_bu$spin_diff / apps_spin_xwoba_bu$pre_spin)
apps_spin_xwoba_bu
```

And I plot the data, with the x axis indicating the percent change in spin rate for a pitcher and the y-axis indicating the xWOBA difference.
```{r}
ggplot(apps_spin_xwoba_bu, aes(x = percent_spin_change , y = xwOBA_diff))+
  geom_point()+
  labs(
    x = "Percent Spin Change",
    y = "xwOBA_diff",
    caption = "Source: Baseball Savant"
  )+
  geom_vline(xintercept = 0.0, linetype="solid", color = "black", size=1)+
  geom_hline(yintercept = 0.0, linetype="solid", color = "black", size=1)
```
It doesn't look like there is a correlation between spin change and xWOBA based on the graph. But, I am going to incorporate the difference in spin rate along with the spin rates before and after June 7th, the change in Bauer Units, and the difference in xWOBA into one graph to see if other variables cause an impact.

```{r}
all_spin_xwoba_corr <- apps_spin_xwoba_bu[,c('xwOBA_diff', 'pitches_per_app', 'pre_spin', 'spin_diff', 'bu_diff')]

cors <- function(df) { 
   # turn all three matrices (r, n, and P into a data frame)
   M <- Hmisc::rcorr(as.matrix(df))
   # return the three data frames in a list return(Mdf)
   Mdf <- map(M, ~data.frame(.x))
}
formatted_cors <- function(df){
 cors(df) %>%
 map(~rownames_to_column(.x, var="measure1")) %>%
 map(~pivot_longer(.x, -measure1, "measure2")) %>% 
 bind_rows(.id = "id") %>%
 pivot_wider(names_from = id, values_from = value) %>%
 mutate(sig_p = ifelse(P < .05, T, F), p_if_sig = ifelse(P <.05, P, NA), r_if_sig = ifelse(P <.05, r, NA)) 
}

formatted_cors(all_spin_xwoba_corr) %>% 
 ggplot(aes(measure1, measure2, fill=r, label=round(r_if_sig,2))) +
 geom_tile() +
 labs(x = NULL, y = NULL, fill = "Pearson's\nCorrelation", title="xwOBA Correlations-All Pitchers", subtitle="Only significant Pearson's correlation coefficients shown") + scale_fill_gradient2(mid="#FBFEF9",low="#0C6291",high="#A63446", limits=c(-1,1)) +
 geom_text() +
 theme_classic() +
 scale_x_discrete(expand=c(0,0)) +
 scale_y_discrete(expand=c(0,0)) 
 


```
However, after looking at the heat map, there doesn't seem to be a strong correlation between xWOBA and other variables as well, which means that spin rate did not have as much of an effect on a player's xWOBA.

Conclusion:
The threat of enforcement of the "sticky stuff" ball rule change had a significant effect on the spin rate and most likely the usage of these substances caused this drop in spin rate by June 7th. However, after its actual enforcement on June 21, the spin rate remained fairly stagnant, indicating that by the time enforcement started, most of the players stopped the use of sticky substances in order to get a better grip. When looking at the distribution of spin rate drops, certain players and teams seemed to have been impacted by this more than others, indicating that some teams were relying on sticky substances more than others. However, the ban of sticky substances did not seem to have an effect on a player's performance in the game, as seen by the lack of correlation between spin rate and xWOBA.The pitchers' xWOBAs remain consistent even with the rule change amongst a variety of factors.

Potential Further Analysis:
Since the data only covers up till the 2021 season, the lack of correlation may change in the upcoming seasons since the rule would then be implemented the entire season rather than just half of a season. As the sample size grows of spin rates and xWOBA beyond June 21, this may change the results. Additionally, I would look at the strikeout rate for these pitchers over time. Maybe the reason their xWOBAs are staying  fairly constant is because their strikeouts are turning into weak contact outs or other types of non-hits.




